---
title: "Somerville K-8 Schools and Students"
output: 
  tint::tintHtml
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
options(java.parameters = "-Xmx2G")

library(tidyverse)
library(sf)
library(tidycensus)
library(tigris)
library(here)
library(ggspatial)
library(ggthemes)
library(ggnewscale)
library(r5r)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
schools <- tibble(school = factor(c("WHCIS (115 Sycamore Street)",
                                    "Hypothetical WHCIS at 8 Bonair Street",
                                    "Hypothetical WHCIS at 554 Broadway",
                             rep("Other SPS school serving K-5 students", 6)),
                             levels = c("WHCIS (115 Sycamore Street)",
                                        "Hypothetical WHCIS at 8 Bonair Street",
                                    "Hypothetical WHCIS at 554 Broadway",
                                        "Other SPS school serving K-5 students")),
                  lat = c(42.391501258379805,
                          42.3876187927341, 
                          42.397961975144824, 
                          42.37905427619121,
                          42.39744670040332, 
                          42.397431825014976, 
                          42.38584202754295, 
                          42.38924933782342, 
                          42.406118297446874),
                  lon = c(-71.09888308341662,
                          -71.08720578602485,
                          -71.10724327216819,
                          -71.09860349311977,
                          -71.09547133152927,
                          -71.11412655504013,
                          -71.08622473784972,
                          -71.11555456799908,
                          -71.12633659736886)) |>
  st_as_sf(coords = c("lon", "lat"), crs = "WGS84")

student_points <- st_read(here("student_points.geojson")) |>
  st_transform("WGS84")
```

This is a project by Carole Turley Voulgaris. The associated
GitHub repository is [here](https://github.com/c-voulgaris/SPS-WH-SRTS). Feel free to
submit an issue with questions or comments.

## Where do school children in Somerville live?

Here is map showing the approximate spatial distribution of Somerville children
enrolled in school (kindergarten through 8th grade) based on data 
from the 2021 5-year sample of the American Community Survey (see this
[technical appendix](https://c-voulgaris.github.io/SPS-WH-SRTS/tech-app.html) 
for details). Each dot represents one student.
Dots are colored to differentiate between children in households with 
incomes above the poverty level and those in households below the
poverty level.

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.cap="Spatial distribution and poverty status of K-8 students in Somerville. Cartolight basemap based on data from Open Street Map (c) OpenStreetMap contributors."}

ggplot(student_points) +
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(color = poverty), size = 0.3, alpha = 0.3) +
  scale_color_manual(values = c("thistle",
                                "salmon2"),
                     labels = c(
          "Student with household income above the poverty level",
          "Student with household income below the poverty level")) +
  new_scale_color() +
  geom_sf(data = schools[schools$school != "Hypothetical WHCIS at 8 Bonair Street" &
                         schools$school != "Hypothetical WHCIS at 554 Broadway",], 
          aes(color = school),
          size = 3) +
  scale_color_manual(values = c("forestgreen", "black")) +
  theme_void() +
  theme(legend.background = element_rect(fill = NA,
                                         linetype = "blank"),
        legend.title = element_blank())

```

This next map is similar, with each dot representing one student and dot 
colors indicating whether the child is non-Hispanic and white
or either non-white or Hispanic^[The American Community Survey records race (white, Black, Asian, etc) and ethnicity (either Hispanic or non-Hispanic) separately.]. 

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Spatial distribution of K-8 students in Somerville, based on block-group-level data from the American Community Survey (2020 5-year sample). Cartolight basemap based on data from Open Street Map (c) OpenStreetMap contributors."}

ggplot(student_points) +
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(color = race), size = 0.3, alpha = 0.3) +
    scale_color_manual(values = c("thistle","salmon2"),
                     labels = c("Non-Hispanic white student",
                                "Non-white or Hispanic student")) +
  new_scale_color() +
  geom_sf(data = schools[schools$school != "Hypothetical WHCIS at 8 Bonair Street" &
                         schools$school != "Hypothetical WHCIS at 554 Broadway",], 
          aes(color = school),
          size = 3) +
  scale_color_manual(values = c("forestgreen", "black")) +
  theme_void() +
  theme(legend.background = element_rect(fill = NA,
                                         linetype = "blank"),
        legend.title = element_blank())

```

## Where is the middle of Somerville?

One way to think about the middle of a shape (for example, the boundary of a
city like Somerville) is its _centroid_. If you cut out a piece of cardboard
in the shape of Somerville, and you wanted to balance your Somerville-shaped
cardboard on the point of a pencil, the centroid is where you'd need to put the
pencil in order for it to balance and stay level.

The _population-weighted centroid_ is another useful way to think about the
middle of a geographic area. If you had that same piece of cardboard, and you 
could stick a bunch of heavy pins to the top of it, with each pin at a student's
home address, and _then_ you tried to balance it on the tip of a pencil, it would
fall off your pencil if you placed it at the centroid you were using before. 
The population-weighted centroid is where
you would have to put the pin to balance your little Somerville cardboard 
now that it's been weighted to reflect the population of students. Another 
term for a population-weighted centroid is the _center of population_.

When we're thinking about where schools are needed, the center of 
the student population is probably more relevant 
than the simple geometric centroid, since land doesn't go to school; children do.

Here is a map showing Somerville's center of the K-8 student population.

```{r, message=FALSE, echo=FALSE, fig.cap="Population-weighted centroid of K-8 students in Somerville. Cartolight basemap based on data from Open Street Map (c) OpenStreetMap contributors."}

bg_pop_center <- as.data.frame(st_coordinates(student_points)) |>
  summarise(lat = mean(Y),
            lon = mean(X)) |>
  st_as_sf(coords = c("lon", "lat"), crs = "WGS84")

ggplot(student_points) +
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(color = "Student enrolled in grade K-8"), size = 0.3, alpha = 0.3) +
  scale_color_manual(values = c("gray50")) +
  new_scale_color() +
  geom_sf(data = schools[schools$school != "Hypothetical WHCIS at 8 Bonair Street" &
                         schools$school != "Hypothetical WHCIS at 554 Broadway",], 
          aes(color = school),
          size = 3) +
  scale_color_manual(values = c("forestgreen", "black")) +
  new_scale_color() +
  geom_sf(data = bg_pop_center,
          size = 5,
          shape = 10,
          aes(color = "Center of student population (grades K-8)")) +
  scale_color_manual(values = "purple") +
  theme_void() +
  theme(legend.background = element_rect(fill = NA,
                                         linetype = "blank"),
        legend.title = element_blank())


```

## Average distances to school

Since Winter Hill Community Innovation School is so centrally-located, 
it plays an important role in minimizing the average distance between
a Somerville student's home and the nearest school. 

Here is a map showing the walking distances^[This is the distance students would need to walk using the existing street network, which is different from the straight-line "as the crow flies" distance. Distances are calculated using the r5r package in R.] to the nearest K-8 
school^[Brown school is included, even though it only serves grades K-5; Capuano is not included, even though it serves Kindergarten and first grade] for 
K-8 students in Somerville.^[Refer to the [technical appendix](https://c-voulgaris.github.io/SPS-WH-SRTS/tech-app.html) for details on how the residential locations of students are approximated.]

```{r, message=FALSE, echo=FALSE, results='hide', warning=FALSE}
points <- student_points |>
  mutate(id = seq(1, nrow(student_points))) 

schools <- schools |>
  mutate(id = seq(1, nrow(schools)))

somer_core <- setup_r5(here("network"))

tt <- travel_time_matrix(somer_core, 
                         origins = points, 
                         destinations = schools)

tt_wide <- tt |> pivot_wider(names_from = toId,
                             values_from = travel_time)

tt_existing <- tt |>
  mutate(id = as.numeric(fromId),
         toId = as.numeric(toId)) |>
  filter(toId != 2, toId != 3) |>
  group_by(id) |>
  mutate(min_time_existing = min(travel_time)) |>
  mutate(closest_school = min_time_existing == travel_time) |>
  filter(closest_school) |>
  group_by(id) |>
  summarise(min_time_existing = mean(travel_time),
            close_school = min(toId))
  
tt_no_wh <- tt |>
  mutate(id = as.numeric(fromId),
         toId = as.numeric(toId)) |>
  filter(toId > 3) |>
  group_by(id) |>
  summarise(min_time_no_wh = min(travel_time)) 
  
tt_edgerly <- tt  |>
  mutate(id = as.numeric(fromId),
         toId = as.numeric(toId)) |>
  filter(toId == 2 | toId > 3) |>
  group_by(id) |>
  summarise(min_time_edgerly = min(travel_time))
  
tt_trum <- tt |>
  mutate(id = as.numeric(fromId),
         toId = as.numeric(toId)) |>
  filter(toId > 2) |>
  group_by(id) |>
  summarise(min_time_trum = min(travel_time)) 

points <- points |>
  left_join(tt_existing) |>
  left_join(tt_no_wh) |>
  left_join(tt_edgerly) |>
  left_join(tt_trum)

stop_r5()
```

```{r, echo=FALSE, message=FALSE, fig.cap="Distances to the nearest school for K-8 students in Somerville (existing school locations). Cartolight basemap based on data from Open Street Map (c) OpenStreetMap contributors."}
ggplot(points) +
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(color = min_time_existing), size = 0.3, alpha = 0.3) +
  scale_color_viridis_c(name = "Walk time to closest school (minutes)") +
  new_scale_color() +
  geom_sf(data = schools[schools$school != "Hypothetical WHCIS at 8 Bonair Street" &
                         schools$school != "Hypothetical WHCIS at 554 Broadway",], 
          aes(color = school),
          size = 3) +
  scale_color_manual(name = "School location", values = c("forestgreen", "black")) +
  new_scale_color() +
  theme_void() +
  theme(legend.background = element_rect(fill = NA,
                                         linetype = "blank"))
```

Here is a map showing the same analysis under a hypothetical condition where
the Winter Hill building is removed from the set of schools without being replaced.

```{r, message=FALSE, echo = FALSE, fig.cap="Distances to the nearest school for K-8 students in Somerville (without WHCIS). Cartolight basemap based on data from Open Street Map (c) OpenStreetMap contributors."}
ggplot(points) +
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(color = min_time_no_wh), size = 0.3, alpha = 0.3) +
  scale_color_viridis_c(name = "Walk time to closest school (minutes)") +
  new_scale_color() +
  geom_sf(data = schools[schools$id > 3,], 
          aes(color = school),
          size = 3) +
  scale_color_manual(name = "School location", values = c("black")) +
  new_scale_color() +
  theme_void() +
  theme(legend.background = element_rect(fill = NA,
                                         linetype = "blank"))
```

Here is the same map for a hypothetical condition where the Winter Hill building
is replaced by a building at 8 Bonair Street.

```{r, message=FALSE, echo=FALSE, fig.cap="Distances to the nearest school for K-8 students in Somerville (WHCIS moved to 8 Bonair Street). Cartolight basemap based on data from Open Street Map (c) OpenStreetMap contributors."}
ggplot(points) +
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(color = min_time_edgerly), size = 0.3, alpha = 0.3) +
  scale_color_viridis_c(name = "Walk time to closest school (minutes)") +
  new_scale_color() +
  geom_sf(data = schools[schools$id > 3 | schools$id == 2,], 
          aes(color = school),
          size = 3) +
  scale_color_manual(name = "School location", values = c("forestgreen", "black")) +
  new_scale_color() +
  theme_void() +
  theme(legend.background = element_rect(fill = NA,
                                         linetype = "blank"))
```

Here is the same map for a hypothetical condition where the Winter Hill building
is replaced by a building at 554 Broadway.

```{r, message=FALSE, echo=FALSE, fig.cap="Distances to the nearest school for K-8 students in Somerville (WHCIS moved to 554 Broadway). Cartolight basemap based on data from Open Street Map (c) OpenStreetMap contributors."}
ggplot(points) +
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(color = min_time_trum), size = 0.3, alpha = 0.3) +
  scale_color_viridis_c(name = "Walk time to closest school (minutes)") +
  new_scale_color() +
  geom_sf(data = schools[schools$id > 2,], 
          aes(color = school),
          size = 3) +
  scale_color_manual(name = "School location", values = c("forestgreen", "black")) +
  new_scale_color() +
  theme_void() +
  theme(legend.background = element_rect(fill = NA,
                                         linetype = "blank"))
```

Obviously, the students for whom the distance to the closest school would
change the most with the removal or relocation of WHCIS would be those
for whom the location at 115 Sycamore Street would have otherwise been the
closest school (shown in the map below).

```{r, echo=FALSE, message=FALSE, fig.cap="Locations of students for whom 115 Sycamore Street is the closest school. Cartolight basemap based on data from Open Street Map (c) OpenStreetMap contributors."}
ggplot(points) +
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(color = close_school == 1), size = 0.3, alpha = 0.3) +
  scale_color_manual(labels = c("Students for whom WHCIS is not the closest school",
                                "Students for whom WHCIS is the closest school"),
                     values = c("gray70",
                                "lightgreen")) +
  new_scale_color() +
  geom_sf(data = schools[schools$school != "Hypothetical WHCIS at 8 Bonair Street" &
                         schools$school != "Hypothetical WHCIS at 554 Broadway",], 
          aes(color = school),
          size = 3) +
  scale_color_manual(values = c("forestgreen", "black")) +
  new_scale_color() +
  theme_void() +
  theme(legend.background = element_rect(fill = NA,
                                         linetype = "blank"),
        legend.title = element_blank())
```


Here is a table showing the average walk time to the closest school
for K-8 students under each of the above scenarios. The first column
shows average walk times to the closest school for all students in Somerville.
The second column shows average walk times to the closest school for those
students for whom the school at 115 Sycamore Street is currently the closest
school.

```{r, echo =FALSE}
tibble(Scenario = c("WHCIS at 115 Sycamore Street",
                    "WHCIS removed without replacement",
                    "WHCIS replaced with school at 8 Bonair Street",
                    "WHCIC replaced with a school at 554 Broadway"),
       `Average walking time to the closest school (all students) (minutes)` = 
         c(mean(points$min_time_existing),
           mean(points$min_time_no_wh),
           mean(points$min_time_edgerly),
           mean(points$min_time_trum)),
       `Average walking time to the closest school (students closest to 115 Sycamore St) (minutes)` = 
         c(mean(points$min_time_existing[points$close_school ==1]),
           mean(points$min_time_no_wh[points$close_school ==1]),
           mean(points$min_time_edgerly[points$close_school ==1]),
           mean(points$min_time_trum[points$close_school ==1]))) |>
  knitr::kable(digits = 1)
```



