---
title: "Technical Appendix"
subtitle: "Somerville Schools and Students"
output: 
  tint::tintHtml
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This analysis was completed using the following packages in the R programming language.

```{r, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(tidycensus)
library(sf)
library(here)
library(tigris)
```


Public data from the American Community Survey includes the number of students
enrolled in school, disaggregated by grade level, at the census block-group level.
Data are further disaggregated by race/ethnicity and, at the tract level, by 
poverty status.

The number of households in each census block is available from the 2020 Census.

I've simulated the residential locations of students by: 

1. Gathering the number of occupied housing units in each census block 
from the 2020 Census
2. Calculating the ratio of non-Hispanic white students to occupied housing units
and the ratio of Hispanic or non-white students to occupied housing units in 
each census block group.
3. Calculating the ratio of students in households above and below the poverty level
to the number of occupied housing units in each census tract.
4. Calculating the number of students in each category for each census block, 
based on the relevant ratio for the tract or block group containing the block.
5. Generate the appropriate number of points in each category for each census block,
randomly distributed within each census block.

## 1. Number of occupied housing units in each census block.

```{r, warning=FALSE, error=FALSE, message=FALSE, results='hide'}
somerville <- places(state = "MA") |>
  dplyr::filter(GEOID == "2562535") 

block_hu <- get_decennial(year = 2020, geography = "block",
                          variables = c("H1_002N"),
                          state = "MA",
                          county = "Middlesex",
                          geometry = TRUE) 

block_centers_somer <- st_centroid(block_hu) |>
  st_filter(somerville)

block_hu_somer <- block_hu |>
  filter(GEOID %in% block_centers_somer$GEOID) |>
  mutate(bg_geoid = substr(GEOID, 1, 12),
         tract_geoid = substr(GEOID, 1, 11)) 
```

## 2. Ratio of non-Hispanic white students and non-white or Hispanic studnets to occupied housing units in each census block group.

```{r, message=FALSE, results='hide'}
bg_vars <- c(nh_white_k = "B14007H_004",
             nh_white_1 = "B14007H_005",
             nh_white_2 = "B14007H_006",
             nh_white_3 = "B14007H_007",
             nh_white_4 = "B14007H_008",
             nh_white_5 = "B14007H_009",
             nh_white_6 = "B14007H_010",
             nh_white_7 = "B14007H_011",
             nh_white_8 = "B14007H_012",
             total_k = "B14007_004",
             total_1 = "B14007_005",
             total_2 = "B14007_006",
             total_3 = "B14007_007",
             total_4 = "B14007_008",
             total_5 = "B14007_009",
             total_6 = "B14007_010",
             total_7 = "B14007_011",
             total_8 = "B14007_012",
             hus = "B25002_002")

bgs <- get_acs(geography = "block group",
               state = "MA",
               county = "Middlesex",
               variables = bg_vars,
               output = "wide",
               year = 2021) |>  
  mutate(nh_white_total = nh_white_kE +
                          nh_white_1E +
                          nh_white_2E +
                          nh_white_3E +
                          nh_white_5E +
                          nh_white_6E +
                          nh_white_7E +
                          nh_white_8E,
         all_total = total_kE +
                     total_1E +
                     total_2E +
                     total_3E +
                     total_5E +
                     total_6E +
                     total_7E +
                     total_8E) |>
  mutate(white_stu_per_hu = nh_white_total / husE,
         non_white_stu_per_hu = (all_total - nh_white_total) / husE) |>
  rename(bg_geoid = GEOID) |>
  select(-NAME)
```

## 2. Ratio of students above and below the poverty level to occupied housing units in each census tract.

```{r, message=FALSE, results='hide'}
tract_vars <- c(pov_gk = "B14006_005",
                pov_g1_4 = "B14006_006",
                pov_g5_8 = "B14006_007",
                no_pov_gk = "B14006_015",
                no_pov_g1_4 = "B14006_016",
                no_pov_g5_8 = "B14006_017",
                hus = "B25002_002")

tracts <- get_acs(geography = "tract",
                  state = "MA",
                  county = "Middlesex",
                  variables = tract_vars,
                  output = "wide",
                  year = 2021) |>  
  mutate(pov_total = pov_gkE +
                     pov_g1_4E +
                     pov_g5_8E,
         no_pov_total = no_pov_gkE +
                        no_pov_g1_4E +
                        no_pov_g5_8E) |>
  mutate(pov_stu_per_hu = pov_total / husE,
         non_pov_stu_per_hu = no_pov_total / husE) |>
  rename(tract_geoid = GEOID) |>
  select(-NAME)
```

## 4. Number of students in each category for each census block

```{r, message=FALSE, results='hide'}
students <- block_hu_somer |>
  left_join(bgs) |>
  mutate(white_students = round(white_stu_per_hu * value, digits = 0),
         non_white_students = round(non_white_stu_per_hu * value, digits = 0)) |>
  rename(block_geoid = GEOID,
         total_hus = value) |>
  left_join(tracts, by="tract_geoid") |>
  mutate(pov_students = round(pov_stu_per_hu * total_hus, digits = 0),
         non_pov_students = round(non_pov_stu_per_hu * total_hus, digits = 0)) |>
  select(block_geoid, bg_geoid, tract_geoid, total_hus, white_students, 
         non_white_students, pov_students, non_pov_students)
```

## 5. Generate points

``` {r, message=FALSE, results='hide'}
student_points_white <- st_sample(students, 
                                    size = students$white_students) |>
   st_as_sf() |>
  mutate(`race` = "Non-Hispanic white student")

student_points_non_white <- st_sample(students, 
                                    size = students$non_white_students) |>
   st_as_sf() |>
  mutate(`race` = "Non-white or Hispanic student")

student_points <- rbind(student_points_white, student_points_non_white) |>
  st_join(students) |>
  mutate(total_students = white_students + non_white_students) |>
  mutate(pct_pov = pov_students / (pov_students + non_pov_students)) |>
  mutate(random = runif(n = n(), min = 0, max = 1)) |>
  mutate(poverty = random < pct_pov) |>
  replace_na(list(poverty = 0)) |>
  select(race, poverty)

st_write(student_points, here("student_points.geojson"), delete_dsn = TRUE)
```

