---
title: "Technical Appendix"
subtitle: "Walking Distances to Somerville Schools"
output: 
  html_document:
    theme: journal
    toc: yes
    toc_float:
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This analysis was completed using the following packages in the R programming language.

```{r, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(tidycensus)
library(sf)
library(here)
library(tigris)
```


Public data from the American Community Survey includes the number of students
enrolled in school, disaggregated by grade level, at the census block-group level.
Data are further disaggregated by race/ethnicity and, at the tract level, by 
poverty status.

The number of households in each census block is available from the 2020 Census.

I've simulated the residential locations of students by: 

1. Gathering the number of occupied housing units in each census block 
from the 2020 Census
2. Calculating the ratio of students to occupied housing units
for each census block group.
3. Calculating the number of students in each each census block, 
based on the relevant ratio for the block group containing the block.
4. Generating the appropriate number of points in each category for each census block,
randomly distributed within each census block.

## 1. Occupied housing units in census blocks.

```{r, warning=FALSE, error=FALSE, message=FALSE, results='hide'}
sf::sf_use_s2(FALSE)

somerville <- places(state = "MA") |>
  dplyr::filter(GEOID == "2562535") 

block_hu <- get_decennial(year = 2020, geography = "block",
                          variables = c("H1_002N"),
                          state = "MA",
                          county = "Middlesex",
                          geometry = TRUE) |>
  erase_water()

block_centers_somer <- st_centroid(block_hu) |>
  st_filter(somerville)

block_hu_somer <- block_hu |>
  filter(GEOID %in% block_centers_somer$GEOID) |>
  mutate(bg_geoid = substr(GEOID, 1, 12),
         tract_geoid = substr(GEOID, 1, 11)) 
```

## 2. Ratio of students to occupied housing units by block group.

```{r, message=FALSE, results='hide'}
bg_vars <- c(grade_k = "B14007_004",
             grade_1 = "B14007_005",
             grade_2 = "B14007_006",
             grade_3 = "B14007_007",
             grade_4 = "B14007_008",
             grade_5 = "B14007_009",
             hus = "B25002_002")

bgs <- get_acs(geography = "block group",
               state = "MA",
               county = "Middlesex",
               variables = bg_vars,
               output = "wide",
               year = 2022) |>  
  mutate(students = grade_kE +
                     grade_1E +
                     grade_2E +
                     grade_3E +
                     grade_4E +
                     grade_5E) |>
  mutate(stu_per_hu = students / husE) |>
  rename(bg_geoid = GEOID) |>
  select(-NAME)
```


## 4. Students by census block

```{r, message=FALSE, results='hide'}
students <- block_hu_somer |>
  left_join(bgs) |>
  mutate(students = round(stu_per_hu * value, digits = 0)) |>
  rename(block_geoid = GEOID,
         total_hus = value) |>
  select(block_geoid, total_hus, students)
```

## 5. Generate points

``` {r, message=FALSE, results='hide', warning=FALSE}
student_points<- st_sample(students, 
                                    size = students$students) |>
   st_as_sf() 

st_write(student_points, here("student-points-2.geojson"), delete_dsn = TRUE)
```

