---
title: "Technical Appendix"
subtitle: "Somerville Schools and Students"
output: 
  tint::tintHtml
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This analysis was completed using the following packages in the R programming language.

```{r, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(tidycensus)
library(sf)
library(here)
library(tigris)
library(leaflet)
```


Public data from the American Community Survey includes the number of students
enrolled in school, disaggregated by grade level, at the census block-group level.

The number of households in each census block is available from the 2020 Census.

I've simulated the residential locations of students by: 

1. Gathering the number of occupied housing units in each census block 
from the 2020 Census
2. Calculating the ratio of students to occupied housing units in 
each census block group.
3. Calculating the number of students for each census block, 
based on the relevant ratio for the tract or block group containing the block.
4. Generate the appropriate number of points in each category for each census block,
randomly distributed within each census block.

## 1. Number of occupied housing units in each census block.

I am filtering the set of census blocks to only include those with at least 
one occupied housing unit.

```{r count-occupied, warning=FALSE, error=FALSE, message=FALSE, results='hide'}
somerville <- places(state = "MA") |>
  dplyr::filter(GEOID == "2562535") 

block_hu <- get_decennial(year = 2020, geography = "block",
                          variables = c("H1_002N"),
                          state = "MA",
                          county = "Middlesex",
                          geometry = TRUE) 

block_centers_somer <- st_centroid(block_hu) |>
  st_filter(somerville)

block_hu_somer <- block_hu |>
  filter(GEOID %in% block_centers_somer$GEOID) |>
  mutate(bg_geoid = substr(GEOID, 1, 12),
         tract_geoid = substr(GEOID, 1, 11))  |>
  filter(value > 0) |>
  st_transform("WGS84")
```

And here is a map showing the number of occupied housing units in each block
with at least one.

```{r map-occupied}
my_pal <- colorNumeric(
  palette = "viridis",
  domain = block_hu_somer$value)

block_labels <- paste0(block_hu_somer$value,
                        " occupied housing units") |>
  lapply(htmltools::HTML)

leaflet(block_hu_somer) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addPolygons(weight = 1,
              color = "white",
              fillColor = ~my_pal(value),
              fillOpacity = 0.3,
              highlightOptions = highlightOptions(weight = 3,
                                                  fillOpacity = 0.5),
              label = block_labels)
```


## 2. Ratio of K-5 students to occupied housing units in each census block group.

```{r pct-students, message=FALSE, warning=FALSE, results='hide'}
bg_vars <- c(total_k = "B14007_004",
             total_1 = "B14007_005",
             total_2 = "B14007_006",
             total_3 = "B14007_007",
             total_4 = "B14007_008",
             total_5 = "B14007_009",
             hus = "B25002_002")

bgs <- get_acs(geography = "block group",
               state = "MA",
               county = "Middlesex",
               variables = bg_vars,
               output = "wide",
               year = 2023,
               geometry = TRUE,
               progress_bar = FALSE) |>  
  mutate(students = total_kE +
                     total_1E +
                     total_2E +
                     total_3E +
                     total_4E +
                     total_5E) |>
  mutate(stu_per_hu = students / husE) |>
  rename(bg_geoid = GEOID) |>
  select(-NAME) 

bg_centers_somer <- st_centroid(bgs) |>
  st_filter(somerville)

bgs_somer <- bgs |>
  filter(bg_geoid %in% bg_centers_somer$bg_geoid) |>
  filter(students > 0) |>
  st_transform("WGS84")
```

Let's see a map of the block groups in which the census estimates that there 
are K-5 students.

```{r map-pct-students}
my_pal <- colorNumeric(
  palette = "viridis",
  domain = bgs_somer$students)

stu_block_labels <- paste0(bgs_somer$husE,
                           " occupied housing units<br/>",
                           bgs_somer$students,
                           " K-5 students (estimated)<br/>",
                           formatC(bgs_somer$stu_per_hu, 
                                   format = "f", 
                                   digits = 2),
                           " estimated K-5 students per<br/>",
                           "         occuplied housing unit") |>
  lapply(htmltools::HTML)

leaflet(bgs_somer) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addPolygons(weight = 1,
              color = "white",
              fillColor = ~my_pal(students),
              fillOpacity = 0.3,
              highlightOptions = highlightOptions(weight = 3,
                                                  fillOpacity = 0.5),
              label = stu_block_labels)
```


## 3. Number of students in each census block

```{r calc-students, message=FALSE, results='hide'}
bgs_somer <- st_drop_geometry(bgs_somer)

students <- block_hu_somer |>
  left_join(bgs_somer) |>
  mutate(students = round(stu_per_hu * value, digits = 0)) |>
  rename(block_geoid = GEOID,
         total_hus = value) |>
  select(block_geoid, 
         bg_geoid, 
         tract_geoid, 
         total_hus, 
         stu_per_hu, 
         students) |>
  filter(students > 0)
```

And here is the distribution of students on a map:

```{r map-students}
my_pal <- colorNumeric(
  palette = "viridis",
  domain = students$total_students)

stu_block_labels <- paste0(students$total_hus,
                           " total occupied housing units<br/>",
                           "(on this block)<br/><br/>",
                           formatC(students$stu_per_hu, 
                                   format = "f", 
                                   digits = 2),
                           " K-5 students per occuplied housing unit<br/>",
                           "(in this block group)<br/><br/>",
                           students$students,
                           " estimated K-5 students on this block") |>
  lapply(htmltools::HTML)

leaflet(students) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addPolygons(weight = 1,
              color = "white",
              fillColor = ~my_pal(students),
              fillOpacity = 0.3,
              highlightOptions = highlightOptions(weight = 3,
                                                  fillOpacity = 0.5),
              label = stu_block_labels)
```

## 4. Generate points

And now we can generate a set of locations that's consistent with the 
distribution in the map above.

``` {r make-pts, message=FALSE, results='hide'}
student_points <- st_sample(students, 
                            size = students$students) |>
   st_as_sf() 

st_write(student_points, here("student_points.geojson"), delete_dsn = TRUE)
```

Here is how those points look on a map.

```{r map-pts}
leaflet(student_points) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addCircles(color = NA,
             radius = 3,
             fillColor = "purple",
             fillOpacity = 0.7)
```


