---
title: "Walking Distances to Somerville Schools"
output: 
  html_document:
    theme: journal
    toc: yes
    toc_float:
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(here)
library(ggspatial)
library(ggthemes)
library(ggnewscale)
library(RColorBrewer)
library(leaflet)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
schools <- tibble(school = factor(c("Winter Hill (or replacement)",
                                    "Winter Hill (or replacement)",
                                    "Winter Hill (or replacement)",
                                    "Brown School",
                                    "Kennedy School", 
                                    "West Somerville School",
                                    "Argenziano School",
                                    "Healey School",
                                    "East Somerville School"),
                             levels = c("Winter Hill (or replacement)",
                                    "Brown School",
                                    "Kennedy School", 
                                    "West Somerville School",
                                    "Argenziano School",
                                    "Healey School",
                                    "East Somerville School")),
                  lat = c(42.391501258379805,
                          42.3876187927341, 
                          42.397961975144824, 
                          42.397431825014976, 
                          42.38924933782342, 
                          42.406118297446874,
                          42.37905427619121,
                          42.39744670040332,
                          42.38584202754295),
                  lon = c(-71.09888308341662,
                          -71.08720578602485,
                          -71.10724327216819,
                          -71.11412655504013,
                          -71.11555456799908,
                          -71.12633659736886,
                          -71.09860349311977,
                          -71.09547133152927,
                          -71.08622473784972),
                  wh_loc = c("Sycamore",
                             "Edgerly",
                             "Trum",
                             rep("none", 6))) |>
  st_as_sf(coords = c("lon", "lat"), crs = "WGS84") 

student_points <- st_read(here("student-walk-times.geojson")) |>
  mutate(close_school_text_prior = factor(case_when(close_school_prior == "1" ~ "Winter Hill (or replacement)",
                                             close_school_prior == "2" ~ "Winter Hill (or replacement)",
                                             close_school_prior == "3" ~ "Winter Hill (or replacement)",
                                             close_school_prior == "4" ~ "Brown School",
                                             close_school_prior == "5" ~ "Kennedy School",
                                             close_school_prior == "6" ~ "West Somerville School",
                                             close_school_prior == "7" ~ "Argenziano School",
                                             close_school_prior == "8" ~ "Healey School",
                                             TRUE ~ "East Somerville School"), 
                                          levels = c("Winter Hill (or replacement)",
                                                     "Brown School", 
                                                     "Kennedy School", 
                                                     "West Somerville School",
                                                     "Argenziano School",
                                                     "Healey School",
                                                     "East Somerville School")),
         close_school_text_now = factor(case_when(close_school_now == "1" ~ "Winter Hill (or replacement)",
                                           close_school_now == "2" ~ "Winter Hill (or replacement)",
                                           close_school_now == "3" ~ "Winter Hill (or replacement)",
                                           close_school_now == "4" ~ "Brown School",
                                           close_school_now == "5" ~ "Kennedy School",
                                           close_school_now == "6" ~ "West Somerville School",
                                           close_school_now == "7" ~ "Argenziano School",
                                           close_school_now == "8" ~ "Healey School",
                                           TRUE ~ "East Somerville School"), 
                                          levels = c("Winter Hill (or replacement)",
                                                     "Brown School", 
                                                     "Kennedy School",
                                                     "West Somerville School",
                                                     "Argenziano School",
                                                     "Healey School",
                                                     "East Somerville School")),
         close_school_text_01_01B = factor(case_when(close_school_01_01B == "1" ~ "Winter Hill (or replacement)",
                                              close_school_01_01B == "2" ~ "Winter Hill (or replacement)",
                                              close_school_01_01B == "3" ~ "Winter Hill (or replacement)",
                                              close_school_01_01B == "4" ~ "Brown School",
                                              close_school_01_01B == "5" ~ "Kennedy School",
                                              close_school_01_01B == "6" ~ "West Somerville School",
                                              close_school_01_01B == "7" ~ "Argenziano School",
                                              close_school_01_01B == "8" ~ "Healey School",
                                              TRUE ~ "East Somerville School"), 
                                          levels = c("Winter Hill (or replacement)",
                                                     "Brown School", 
                                                     "Kennedy School", 
                                                     "West Somerville School",
                                                     "Argenziano School",
                                                     "Healey School",
                                                     "East Somerville School")),
         close_school_text_04 = factor(case_when(close_school_04 == "1" ~ "Winter Hill (or replacement)",
                                          close_school_04 == "2" ~ "Winter Hill (or replacement)",
                                          close_school_04 == "3" ~ "Winter Hill (or replacement)",
                                          close_school_04 == "4" ~ "Brown School",
                                          close_school_04 == "5" ~ "Kennedy School",
                                          close_school_04 == "6" ~ "West Somerville School",
                                          close_school_04 == "7" ~ "Argenziano School",
                                          close_school_04 == "8" ~ "Healey School",
                                          TRUE ~ "East Somerville School"), 
                                          levels = c("Winter Hill (or replacement)",
                                                     "Brown School", 
                                                     "Kennedy School", 
                                                     "West Somerville School",
                                                     "Argenziano School",
                                                     "Healey School",
                                                     "East Somerville School"))) 
```

This is a project by Carole Turley Voulgaris. The associated
GitHub repository is [here](https://github.com/c-voulgaris/SPS-WH-SRTS). Feel free to
submit an issue with questions or comments.

## Children's closest schools {.tabset}

The tabs below include maps showing the approximate spatial distribution of 
Somerville children enrolled in school (kindergarten through 5th grade) based on data 
from the 2023 5-year sample (2019-2023) of the American Community Survey (see this
[technical appendix](https://c-voulgaris.github.io/SPS-WH-SRTS/tech-app.html) 
for details). Each small dot represents one student, and the larger dots 
represent schools. There are some areas where no students appear. These are locations
where there K-5 students were found in the American Community Survey between
2019 and 2023, so the estimated number of students in those areas is zero, although
there is a margin of error for those estimates, so there could be students living
in those areas. A more precise analysis could use actual students addresses, but 
those locations are not publicly available (and for good reason). 



The dots representing students are color-coded to indicate which school a student
lives closest to (in terms of walking distance) under each scenario. 

### Pre-closure

```{r, message=FALSE, warning=FALSE, echo=FALSE}
school_pal <- colorFactor(brewer.pal(7, "Dark2"), schools$school)

these_schools <- schools[schools$wh_loc != "Trum" &
                         schools$wh_loc != "Edgerly",]

leaflet(student_points) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addCircles(color = NA,
             radius = 3,
             fillColor = ~school_pal(student_points$close_school_text_prior),
             fillOpacity = 0.7) |>
  addCircleMarkers(data = these_schools,
             radius = 5,
             color = "black",
             fillColor = ~school_pal(these_schools$school),
             label = these_schools$school) |>
  addLegend(position = "bottomleft",
            pal = school_pal,
            values = ~these_schools$school,
            title = "Closest school")
```

The above map shows the closest school for each student prior to the closure of 
building at 115 Sycamore St in spring of 2023. 
At that time, Winter Hill School was the closest school for about 23 percent of 
all K-5 students living in Somerville and Brown School was the closest for about 
9 percent of all Somerville K-5 students.

The city could choose to renovate both the Winter Will and Brown schools at their 
existing locations, so students' closest schools would be the same as before
the closure of the Winter Hill building at 115 Sycamore St.

### Existing

```{r, message=FALSE, warning=FALSE, echo=FALSE}
these_schools <- schools[schools$wh_loc != "Trum" &
                         schools$wh_loc != "Sycamore",]

leaflet(student_points) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addCircles(color = NA,
             radius = 3,
             fillColor = ~school_pal(student_points$close_school_text_now),
             fillOpacity = 0.7) |>
  addCircleMarkers(data = these_schools,
             radius = 5,
             color = "black",
             fillColor = ~school_pal(these_schools$school),
             label = these_schools$school) |>
  addLegend(position = "bottomleft",
            pal = school_pal,
            values = ~these_schools$school,
            title = "Closest school")
```


With the closure of the building at 115 Sycamore St and the relocation of
Winter Hill School students to the Edgerly Education Center, the closest school 
for most students who were previously closest to Winter Hill became Argenziano
or Healey

### Brown closure

```{r, message=FALSE, warning=FALSE, echo=FALSE}
these_schools <- schools[schools$wh_loc != "Trum" &
                         schools$wh_loc != "Edgerly" &
                         schools$school != "Brown School",]

leaflet(student_points) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addCircles(color = NA,
             radius = 3,
             fillColor = ~school_pal(student_points$close_school_text_01_01B),
             fillOpacity = 0.7) |>
  addCircleMarkers(data = these_schools,
             radius = 5,
             color = "black",
             fillColor = ~school_pal(these_schools$school),
             label = these_schools$school) |>
  addLegend(position = "bottomleft",
            pal = school_pal,
            values = ~these_schools$school,
            title = "Closest school")
```

This map shows students' closest schools under any option that involves rebuilding 
the Winter Hill school on Sycamore Street, leaving the Brown school building 
vacant and adding capacity at the Kennedy school or at Winter Hill to replace the 
lost capacity at Brown.

These proposals differ in terms of _where_ capacity would be added, and hence 
the number of students who would be able to attend their closest school, but 
they are identical in terms of which school a student would be closest to.

Most students for whom Brown was previously the closest school would be 
closest to West Somerville or Kennedy under these options.

### Trum Field

```{r, message=FALSE, warning=FALSE, echo=FALSE}
these_schools <- schools[schools$wh_loc != "Sycamore" &
                         schools$wh_loc != "Edgerly" &
                         schools$school != "Brown School",]

leaflet(student_points) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addCircles(color = NA,
             radius = 3,
             fillColor = ~school_pal(student_points$close_school_text_04),
             fillOpacity = 0.7) |>
  addCircleMarkers(data = these_schools,
             radius = 5,
             color = "black",
             fillColor = ~school_pal(these_schools$school),
             label = these_schools$school) |>
  addLegend(position = "bottomleft",
            pal = school_pal,
            values = ~these_schools$school,
            title = "Closest school")
```

Another alternative would be to leave both the Winter Hill and Brown school 
buildings vacant and
build a new school at Trum Field. This new school would be the closest school
for about 10 percent of all K-5 students in Somerville, including about 21 percent
of the students for whom Winter Hill was previously the closest school and
about 52 percent of the students from whom Brown was previously the closest 
school.

## Walk time maps {.tabset}

The following maps show illustrate walking times^[Travel times are calculated using the r5r package in R.] to the nearest school that serves students in at least grades kindergarten through fifth grade.^[Refer to the [technical appendix](https://c-voulgaris.github.io/SPS-WH-SRTS/tech-app.html) for details on how the residential locations of students are approximated.]

### Pre-closure

```{r, message=FALSE, warning=FALSE, echo=FALSE}
dist_pal <- colorNumeric("viridis", 
                         student_points$min_time_04,
                         reverse = TRUE)

these_schools <- schools[schools$wh_loc != "Trum" &
                         schools$wh_loc != "Edgerly",]

leaflet(student_points) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addCircles(color = NA,
             radius = 3,
             fillColor = ~dist_pal(student_points$min_time_prior),
             fillOpacity = 0.7) |>
  addCircleMarkers(data = these_schools,
             radius = 5,
             color = "black",
             fillColor = ~school_pal(these_schools$school),
             label = these_schools$school) |>
  addLegend(position = "bottomleft",
            pal = dist_pal,
            values = ~student_points$min_time_prior,
            title = "Walk time to closest school")
```

The above map illustrates walking distances for students prior to the closure of 
building at 115 Sycamore St in spring of 2023. 

Rebuilding/renovating both Winter Will and Brown schools at their 
existing locations, would provide the same walking distances as this prior
condition, since all schools would be in their original locations.

### Existing

```{r, message=FALSE, warning=FALSE, echo=FALSE}
these_schools <- schools[schools$wh_loc != "Trum" &
                         schools$wh_loc != "Sycamore",]

leaflet(student_points) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addCircles(color = NA,
             radius = 3,
             fillColor = ~dist_pal(student_points$min_time_now),
             fillOpacity = 0.7) |>
  addCircleMarkers(data = these_schools,
             radius = 5,
             color = "black",
             fillColor = ~school_pal(these_schools$school),
             label = these_schools$school) |>
  addLegend(position = "bottomleft",
            pal = dist_pal,
            values = ~student_points$min_time_prior,
            title = "Walk time to closest school")
```

This map illustrates walking times for students to walk to the nearest
school now, with the Winter Hill School moved to Edgerly. 
It is worth noting that these are _not_ the walking distances that 
students experience, but rather the distance to the _closest_ school.

Most Winter Hill students who were moved to Edgerly when the location
at 115 Sycamore St closed are presumably not attending their closest school.

### Brown closure

```{r, message=FALSE, warning=FALSE, echo=FALSE}
these_schools <- schools[schools$wh_loc != "Trum" &
                         schools$wh_loc != "Edgerly" &
                           schools$school != "Brown School",]

leaflet(student_points) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addCircles(color = NA,
             radius = 3,
             fillColor = ~dist_pal(student_points$min_time_01_01B),
             fillOpacity = 0.7) |>
  addCircleMarkers(data = these_schools,
             radius = 5,
             color = "black",
             fillColor = ~school_pal(these_schools$school),
             label = these_schools$school) |>
  addLegend(position = "bottomleft",
            pal = dist_pal,
            values = ~student_points$min_time_prior,
            title = "Walk time to closest school")
```

This map illustrates walking times to the nearest school under any option that 
involves rebuilding the Winter Hill school on 
its current site, leaving the Brown school building vacant and adding capacity at the 
Kennedy school or at Winter Hill to replace the lost capacity at Brown.

These proposals differ in terms of _where_ capacity would be added, and hence 
the number of students who would be able to attend their closest school, but 
they are identical in terms of the distance each student would be from the 
_closest_ school, even if their closest school is not their assigned school.

### Trum Field

```{r, message=FALSE, warning=FALSE, echo=FALSE}
these_schools <- schools[schools$wh_loc != "Sycamore" &
                         schools$wh_loc != "Edgerly" &
                           schools$school != "Brown School",]

leaflet(student_points) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addCircles(color = NA,
             radius = 3,
             fillColor = ~dist_pal(student_points$min_time_04),
             fillOpacity = 0.7) |>
  addCircleMarkers(data = these_schools,
             radius = 5,
             color = "black",
             fillColor = ~school_pal(these_schools$school),
             label = these_schools$school) |>
  addLegend(position = "bottomleft",
            pal = dist_pal,
            values = ~student_points$min_time_prior,
            title = "Walk time to closest school")
```

This map illustrates walking times to the nearest school under Option 04,
which would involve leaving both the Brown school the Winter Hill
school at 115 Sycamore St vacant. A new school would be built at Trum
field to accommodate the lost capacity from both schools.

## Summary statistics {.tabset}

```{r, echo=FALSE}

```

The decision to leave two centrally-located school vacant would have important
implications for the time required for Somerville students to walk to school.
This is particularly true for students for whom Brown or Winter Hill would
otherwise have been the closest school.

```{r, echo =FALSE}
summary <- tibble(Scenario = factor(rep(c("Pre-closure",
                        "Existing",
                        "Brown closure",
                        "Trum Field"),3),
                        levels = c("Pre-closure",
                                   "Existing",
                                   "Brown closure",
                                   "Trum Field")),
       Population = c(rep("All K-5\nstudents in\nSomerville",4),
                      rep("K-5 students\nfor whom\n115 Sycamore St\nis the closest school",4),
                      rep("K-5 students\nfor whom Brown is\nthe closest school",4)),
       `Average walking time to the closest school (minutes)` =
         c(mean(student_points$min_time_prior),
           mean(student_points$min_time_now),
           mean(student_points$min_time_01_01B),
           mean(student_points$min_time_04),
           mean(student_points$min_time_prior[student_points$close_school_prior ==1]),
           mean(student_points$min_time_now[student_points$close_school_prior ==1]),
           mean(student_points$min_time_01_01B[student_points$close_school_prior ==1]),
           mean(student_points$min_time_04[student_points$close_school_prior ==1]),
           mean(student_points$min_time_prior[student_points$close_school_prior ==4]),
           mean(student_points$min_time_now[student_points$close_school_prior ==4]),
           mean(student_points$min_time_01_01B[student_points$close_school_prior ==4]),
           mean(student_points$min_time_04[student_points$close_school_prior ==4])),
       `Number of students with less than a\n10-minute walk to the closest school` = 
         c(sum(student_points$min_time_prior < 11),
           sum(student_points$min_time_now < 11),
           sum(student_points$min_time_01_01B < 11),
           sum(student_points$min_time_04 < 11),
           sum(student_points$min_time_prior[student_points$close_school_prior ==1]<11),
           sum(student_points$min_time_now[student_points$close_school_prior ==1]<11),
           sum(student_points$min_time_01_01B[student_points$close_school_prior ==1]<11),
           sum(student_points$min_time_04[student_points$close_school_prior ==1]<11),
           sum(student_points$min_time_prior[student_points$close_school_prior ==4]<11),
           sum(student_points$min_time_now[student_points$close_school_prior ==4]<11),
           sum(student_points$min_time_01_01B[student_points$close_school_prior ==4]<11),
           sum(student_points$min_time_04[student_points$close_school_prior ==4]<11)),
       `Number of students with more than a\n20-minute walk to the closest school` = 
         c(sum(student_points$min_time_prior > 20),
           sum(student_points$min_time_now > 20),
           sum(student_points$min_time_01_01B > 20),
           sum(student_points$min_time_04 > 20),
           sum(student_points$min_time_prior[student_points$close_school_prior ==1]> 20),
           sum(student_points$min_time_now[student_points$close_school_prior ==1]> 20),
           sum(student_points$min_time_01_01B[student_points$close_school_prior ==1]> 20),
           sum(student_points$min_time_04[student_points$close_school_prior ==1]> 20),
           sum(student_points$min_time_prior[student_points$close_school_prior ==4]> 20),
           sum(student_points$min_time_now[student_points$close_school_prior ==4]> 20),
           sum(student_points$min_time_01_01B[student_points$close_school_prior ==4]> 20),
           sum(student_points$min_time_04[student_points$close_school_prior ==4]> 20))) 
```


### Average walks

```{r, echo=FALSE}
ggplot(summary) +
  geom_bar(aes(x = Population,
               y = `Average walking time to the closest school (minutes)`,
               fill = Scenario),
               stat = "identity",
               position = "dodge") +
  scale_fill_brewer(type = "qual") +
  theme_minimal()

```

With the distribution of schools in Somerville prior to the closure of Winter Hill,
the average time it would take a K-5 student in Somerville to walk to the nearest 
school was about 11 minutes. This has increased with the closure of Winter Hill. 
All of the options under consideration, with the exception of the 01+01A (which
would keep all schools at their prior locations), would increase average walking
distances relative to the period before Winter Hill closed.

Average walking times for students who are currently closest to Brown or Winter
Hill would increase by more than 50 percent in scenarios where their respective
schools are left vacant.

### 10-min walks

```{r, echo=FALSE}
ggplot(summary) +
  geom_bar(aes(x = Population,
               y = `Number of students with less than a\n10-minute walk to the closest school`,
               fill = Scenario),
               stat = "identity",
               position = "dodge") +
  scale_fill_brewer(type = "qual") +
  theme_minimal()
```

Just under 1200 K-5 students in Somerville (about 45 percent of all K-5 students) 
lived within a ten-minute walk of a school before the closure of Winter Hill. 
Winter Hill School was the closest school for more than a quarter of all students
within ten minutes of a school, and Brown was the closest school for just over ten 
percent.

Now that the Winter Hill School has closed, only about 20 students for whom the \
Winter Hill School at 115 Sycamore St was the closest school live within ten 
minutes of a school. This number would increase to about 60 students if a new 
School were built at Trum Field, which is about a fifth of the number who live
within 10 minutes of 115 Sycamore St.

A negligible number of students (fewer than ten) for whom Brown is currently the
closest school would be able to walk to any school within ten minutes if the 
Brown school were left vacant and no new school were built. This number would increase
to about 45 students if a new school were built at Trum Field.

### 20-min walks

```{r, echo=FALSE}
ggplot(summary) +
  geom_bar(aes(x = Population,
               y = `Number of students with more than a\n20-minute walk to the closest school`,
               fill = Scenario),
               stat = "identity",
               position = "dodge") +
  scale_fill_brewer(type = "qual") +
  theme_minimal()
```

Prior to the closure of the school at 115 Sycamore St, only about three dozen
K-5 students in Somerville lived more than a 20-minute walk from a school serving 
K-5 students. The closure of Winter Hill School has increased this number by
about 270 percent.
